\jolideco: Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-01-31 src/tex/ms.tex
[showyourwork] overleaf sync

    \section{Method}
    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple
    low counts observations.
    Most generally we assume our total dataset consists of $M$ individual
    observations of the same region of the sky, which are jointly
    modelled. The assumption is that the underlying \textit{true} emission image,
    we are looking for does not change with time. Under this assumption these
    datasets can be for example:

@@

    Or a even an arbitrary combination of the possibilities listed above.

    For each individual observation $m$ the predicted counts can be modelled by forward
    folding the unknown flux image $\mathbf{x}$ with the individual (per observation) instrument response:

    \begin{equation}
        \label{eq:model}
        \mathbf{\lambda}_m = \mathrm{PSF}_m \circledast \left(\mathbf{E}_m \cdot (\mathbf{x} + \mathbf{B}_m) \right)
    \end{equation}

    Where the expected counts $\lambda_i$ are given by the convolution of the true underlying
    flux distribution $\mathbf{x}$ with the $\mathrm{PSF}$. Additionally
    the observation specific exposure $\mathbf{E}$ and background emission $\mathbf{B}$ can be
    taken into account.

    Given a single observation $m$ of a true underlying flux distribution
    $\mathbf{x}$ and assuming the noise in each pixel $i$ in the recorded counts image
    $\mathbf{d}$ follows a Poisson distribution, the total likelihood $\mathcal{L}_m$
    of obtaining the measured image from a model image of the expected
    counts $\lambda_i$ with $N$ pixels is given by:

    \begin{equation}
        \label{eq:poisson}
        \mathcal{L}_m\left( \mathbf{d} | \mathbf{\lambda} \right) = \prod_i^N \frac{{e^{ - d_i } \lambda_i ^ {d_i}}}{{d_i!}}
    \end{equation}

@@

    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{d} | \mathbf{\lambda} \right) = \sum_i^N \lambda_i - d_i \log{\lambda_i}
    \end{equation}

@@

    \begin{equation}
        \label{eq:bayes}
        P(\theta|\textbf{D}) = P(\theta ) \frac{\mathcal{C}(\textbf{D} |\theta)}{P(\textbf{D})}
    \end{equation}

    The total objective function $\mathcal{L}$ is given by:
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left( \mathbf{d_m} | \mathbf{x} \right) = \sum_m^M \mathcal{C}\left( \mathbf{d_m} | \mathbf{x} \right) - \beta \cdot \mathcal{P}(\mathbf{x})
    \end{equation}

@@

    \subsection{Calibration of Predicted Counts}
    Each dataset comes with systematic errors on absolute position and background normalisation

    \begin{itemize}
        \item allow for systematic shifts between observations 
        \item allow for background scaling per observation $\alpha_m$
    \end{itemize}

    \begin{equation}
        \label{eq:model-npred-calibration}
        \mathbf{\lambda}_m = \mathrm{PSF}_m \circledast \left(\mathbf{E}_m \cdot (\phi_m(\mathbf{x}| \delta_x, \delta_y) + \alpha_m \cdot \mathbf{B}_m) \right)
    \end{equation}

    Show example residual images before and after calibration...

@@

    The \jolideco method was implemented as an independent Python package, 
    based on \texttt{PyTorch} as optimization back-end. We tried to define a modular,
    object oriented code structure, to allow parts of the algorithm to be
    flexible and interchangeable, such as choice of image normalisation scales,
    the choice of GMM models, optimization methods and serilisation formats for the 
    reconstruction results as well as corresponding diagnostic information
    such as the trace of the posterior and prior likelihood values.

    The package is available at \url{https://github.com/jolideco/jolideco}

@@

    In addition we use \texttt{Scikit-Learn}~\citep{Skimage2014} to learn the GMM.

    We use \texttt{Astropy}~\citep{Astropy2018} for FITS serialisation and
    handling WCS transforms.

    We use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and
    \texttt{Matplotlib}~\citep{Hunter2007} for plotting.

    For optimization and internal data handling we use \texttt{PyTorch}~\citep{Pytorch2019}.

@@

>>>>

    \section{Method}
    \subsection{Poisson Joint Likelihood}
    Our goal is to recover an image $\mathbf{x}$ from a set of multiple
    low counts observations.
    Most generally we assume our total dataset consists of $J$ individual
    observations of the same region of the sky, which are jointly
    modelled. The assumption is that the underlying \textit{true} emission image,
    we are looking for does not change with time. Under this assumption these
    datasets can be for example:

@@

    Or even an arbitrary combination of the possibilities listed above.

    For each individual observation $j$ the predicted counts can be modelled by forward
    folding the unknown flux image $\mathbf{x}$ with the individual (per observation) instrument response:

    \begin{equation}
        \label{eq:model}
        \boldsymbol{\lambda}_j = \mathrm{PSF}_j \circledast \left(\mathbf{E}_j \cdot (\mathbf{x} + \mathbf{B}_j) \right)
    \end{equation}

    Where the expected counts $\lambda_j$ are given by the convolution of the unknown 
    flux image $\mathbf{x}$ with the observation specific point spread function $\mathrm{PSF}_j$. Additionally
    the observation specific exposure $\mathbf{E}_j$ and background emission $\mathbf{B_j}$ can be
    taken into account.

    Given a single observation $j$ of an unknown flux image
    $\mathbf{x}$ and assuming the noise in each pixel $i$ in the recorded counts image
    $\mathbf{D_j}$ follows a Poisson distribution, the likelihood $\ell_j$
    of obtaining the measured image from a model image of the expected
    counts $\boldsymbol{\lambda_j}$ with $N$ pixels is given by:

    \begin{equation}
        \label{eq:poisson}
        \ell\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \prod_{i=1}^N \frac{{e^{ - D_{j,i} } \lambda_i ^ {D_{j,i}}}}{{D_{j,i}!}}
    \end{equation}

@@

    \begin{equation}
        \label{eq:cash}
        \mathcal{C}\left( \mathbf{D_j} | \boldsymbol{\lambda_j} \right) = \sum_{i=1}^N \lambda_{j,i} - D_{j, i} \log{\lambda_{j,i}}
    \end{equation}

    By summing over all observations we finally get the joint log-likelihood of measuring a set of counts images $\mathbf{D}$ given a unknown flux image $\mathbf{x}$:

    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left( \mathbf{D} | \mathbf{x} \right) = \sum_{j=1}^J \mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)
    \end{equation}

    Using the Maximum Likelihood Estimation (MLE) we could estimate

    \todo{this sum corresponds to the loop over datasets in machine learning... }

@@

    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = \mathcal{P}(\mathbf{x} ) \frac{\mathcal{L}(\textbf{D} |\mathbf{x})}{ \mathcal{P} (\textbf{D})}
    \end{equation}

    The total objective function $\mathcal{L}$ is given by:
    \begin{equation}
        \label{eq:total}
        \mathcal{L}\left( \mathbf{d} | \mathbf{x} \right) = \sum_m^M \mathcal{C}\left( \mathbf{d_m} | \mathbf{x} \right) - \beta \cdot \mathcal{P}(\mathbf{x})
    \end{equation}

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each measured counts image $\mathbf{D_j}$ is associated with systematic errors.
    This includes e.g. the absolute position of the sources in the image and the
    normalisation of the background image, which can vary from observation to observation.
    To account for this we extend the model for the predicted
    counts introduced in Eq.~\ref{eq:model}, by a background normalisation factor
    $\alpha_j$ and a linear shift in position $\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j})$,
    which is specific for each observation. This gives the following modified model definition:

    \begin{equation}
        \label{eq:model-npred-calibration}
        \mathbf{\lambda}_j = \mathrm{PSF}_j \circledast \left(\mathbf{E}_j \cdot (\phi_j(\mathbf{x}| \delta_{x,j}, \delta_{y,j}) + \alpha_j \cdot \mathbf{B}_j) \right)
    \end{equation}

    In practice one observation is used as a reference and its corresponding
    systematic shifts in position $\delta_{x,j}, \delta_{y,j}$ are frozen during
    optimization.

@@

    The \jolideco method was implemented as an independent Python package, 
    based on \texttt{PyTorch} as optimization back-end. We tried to define a modular,
    object oriented code structure, to allow parts of the algorithm to be
    flexible and interchangeable, such as choice of image normalisation scales,
    the choice of GMM models, optimization methods and serialisation formats for the 
    reconstruction results as well as corresponding diagnostic information
    such as the trace of the posterior and prior and likelihood values.
    The package is available at \url{https://github.com/jolideco/jolideco}

@@

    We also use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and
    \texttt{Matplotlib}~\citep{Hunter2007} for plotting.
    To learn the Gaussian Mixture Models from patches we the standard implementation
    provided in the \texttt{Scikit-Learn}~\citep{Skimage2014} package. To handle the 
    serialisation of images to the FITS data format and to handle world coordinate
    systems (WCS) we use \texttt{Astropy}~\citep{Astropy2018}.
    For optimization and internal data handling we use \texttt{PyTorch}~\citep{Pytorch2019}.

@@

    \subsection{Optimization Strategy}

    \begin{itemize}
        \item First run with uniform prior, to get a rough estimate
        \item First optimize the systematic error parameters
        \item The go for the optimization uisng the patch prior, which can be computationally costly
    \end{itemize}
